### 2004-07-16 Documentation/part1.tex
part 1 finis

In this first chapter, we show a \GLOBES\ tour illustrating the
main features of \GLOBES . The complete example  
can be found as {\tt example-tour.c} in the example subdirectory of your \GLOBES\ distribution.
The output is written to {\tt stream}, which can be either {\tt stdout},
or a file. Details about how to use \GLOBES\ with C can found in \Chapt~\ref{chapt:gettingstarted} and the following chapters.
You can also find a summary of the most important \GLOBES\ $\chi^2$-functions in \tabl{stdfunctions}. Note that this chapter
can be skipped without loss of relevant information.

@@

 for(i=1;i<4;i++) fprintf(stream,"1->\%i: \%g ",i,\\ \hspace*{4cm} glbProfileProbability(1,i,+1,50)); \\

@@

we have to destroy them first:

@@

Finally, we have to destroy the parameter vectors again:

@@

\chapter{\GLOBES\ basics}
\label{chapt:gettingstarted}

\index{norm}{Installation}
In this first chapter of the user's manual, we assume that the \GLOBES\ software is readily installed on your computer system. For the installation,
see \App~\ref{app:installation} and the {\tt INSTALL} file in the
software package. We demonstrate how to load pre-defined experiments 
and introduce the basic concepts of \GLOBES . We do not go
into details of the programming language, which means that standard parts
of the program code common to all of the examples in the following chapters are, in general, omitted.
An example of a minimal \GLOBES\ program in C can be found on page~\pageref{ex:c}. Furthermore, the files of the examples in this part can be found in the {\tt Example} subdirectory of your \GLOBES\ distribution. \index{norm}{Examples} Since it depends on your installed
configuration, we refer to the {\tt README} file in the \GLOBES\ main directory, and the comments in the Makefile of the {\tt Example} subdirectory for how to compile the example files. Make sure that
the data files (\AEDL\ and supporting files) in the {\tt data} 
subdirectory can be found by the
examples. In case of doubt, you can simply copy the necessary files
into the working directory where the examples are executed.

@@

\hspace*{0.5cm}  /* Destroy parameter vector(s) */ \\
\hspace*{0.5cm}  glbFreeParams(true\_values); \\
\hspace*{0.5cm} /* ... */ \\

@@

\index{norm}{Experiment!number of}
In principle, the \GLOBES\ user interface can currently handle up to 32 of different long-baseline experiments simultaneously, where the number
of existing experiment definition files can, of course, be unlimited. This means that their $\Delta \chi^2$-values are added {\em after} the minimization over the systematics parameters, and {\em before} any minimization over the oscillation parameters. Note that each experiment
assumes a specific matter density profile, which means
that it makes sense to simulate different operation modes within one
experiment definition, and physically different baselines in different
definitions. For details of the rate computation and
simulation techniques, we refer at this place to \Part~\ref{part:2}. Though
 the simplest case of simulating one experiment may be most often used, 
 using more than one experiments are useful in many cases. For example, combinations of experiments can be tested for
complementarity and competitiveness by equal means within one program.
In general, many \GLOBES\ functions take the experiment number as
a parameter, which runs from $0$ to \GLB{\_num\_of\_exps}-1 in the order of their initialization in the program.\footnote{Note that
the global variable {\tt glb\_num\_of\_exps} must not be modified by the
user.} In addition, using the parameter value \GLBC{GLB\_ALL} as
experiment number initiates a combined analysis of all loaded experiments.

\index{norm}{Experiment!list}
\index{norm}{Initialization!experiments}
For storing the experiments, \GLOBES\ uses the initially empty list of experiments \GLB{\_experiment\_list}. To add a pre-defined experiment to this list, one can use the function {\tt glbInitExperiment}:
\begin{function}
\index{norm}{Experiment initialization}
\GLBNS{InitExperiment}
{\tt int glbInitExperiment(char *inf, glb\_exp *in, int *counter)}
 adds a single experiment with the filename {\tt inf} to the list of currently loaded experiments. The {\tt counter} is a pointer to the 
 variable containing the number of experiments, and the experiment {\tt in}
 points to the beginning of the experiment list. The function returns
 zero if it was successful. 
\end{function}
Normally, a typical  call of {\tt glbInitExperiment} is 
\begin{quote}
{\tt glbInitExperiment("NuFact.glb",\&glb\_experiment\_list[0],\\  \hspace*{8cm} \&glb\_num\_of\_exps); }
\end{quote}
In this case, the experiment in the file {\tt NuFact.glb} is added to the internal global list of experiments, and the experiment counter is increased. The
experiment then has the number {\tt glb\_num\_of\_exps}-1. The elements
of the experiment list have the type \GLB{\_exp}, which the
user will not need to access directly in most cases. The experiment definition files, which usually end with {\tt .glb}, and any
supporting files, are first of all searched in the current directory, and then in the global path variable 
%
\index{norm}{GLB PATH@{\tt GLB\_PATH}}
\index{norm}{Environment variables!GLB PATH@{\tt GLB\_PATH}}
%
set in your environment.
\index{norm}{Path resolution}
A list of pre-defined experiment prototypes, their filenames, their short descriptions, and the references of their definitions can be found in \tabl{experiments}. If the program cannot find these files, or some of them are syntactically not correct, it will break at this place.

@@

Before we can set the simulated event rates or access any oscillation
parameters, we need to become familiar with the concept \GLOBES\ uses
for oscillation parameters. In order to transfer sets of oscillation
parameter vectors $(\theta_{12}, \theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm)$, the parameter type \GLB{\_params\_type} is used. In general, pointers to this structure are of the type \GLB{\_params}, and are often
transferred to and from \GLOBES\ functions. Therefore, the memory space
for these vectors has to be reserved before they can be used, and it has
to be returned afterwards. \GLOBES\ functions usually use the pointers
of the type {\tt glb\_params} for the input or output to the functions. As
an input parameter, the pointer has to be created and point towards a valid
parameter structure, where the oscillation parameters are read from. As
an output parameter, the pointer has to be created, too, and point towards
a structure which will contain the return values will be written to.
This parameter transfer concept seems to be very sophisticated, but, as we
will see in the next chapters, it hides a lot of complicated parameter 
mappings which otherwise need to be done by the user. For example, not
only the oscillation parameters are stored in the pointer structure, but also information on the matter
densities of all of the initialized experiments. Since \GLOBES\ treats
the matter density as a free parameter known with some external precision
to include matter density uncertainties,
the minimizers also use fit values and external errors for the matter
densities of all loaded experiments. More precisely, the matter density
profile of each experiment $i$ is multiplied by a scaling factor
$\hat{\rho_i}$, \index{norm}{Matter density scaling factor}
\index{norm}{Matter density profile} 
which is stored in the density information of {\tt glb\_params}. 
Each of
these scaling factors has $1.0$ as pre-defined value. Since it
is in most cases not necessary to change this value, the user does
not need to take care of it. For a constant matter density, it is simply the ratio of the matter density and the average matter density specified in the experiment definition, \ie , $\hat{\rho_i} \equiv \rho_i/\bar{\rho}_i$. For a matter density profile, it acts as an overall normalization factor: The matter density in each layer is multiplied by this factor. In most cases one wants to take a scaling factor of $1.0$ here, which simply means taking the matter density profile as it is given in the experiment definition. For the treatment of correlations, however,
an external precision of the scaling factor might be used to include the
correlations with the matter density uncertainty. For more details, we
refer to the following sections.
Note that the {\tt glb\_params}
structures must not be initialized before all experiments are loaded,
since the number of matter densities can only be determined after the
experiments are initialized. 
Similarly, any change in the number of experiments requires that the
parameter structures be re-initialized, \ie, destroyed and created again.

Another piece of information will be returned from the minimizers 
(\cf, \Chapt~\ref{chapt:correlations}) and transferred
into the {\tt glb\_params} structure is the number
of iterations used for the minimization, which is proportional to the
running time of the minimizer. In general, the user does not need to
access the elements in {\tt glb\_params} directly. A number of functions
is provided to handle these parameter structures:
\begin{function}
\GLBNS{AllocParams}
{\tt glb\_params glbAllocParams()} allocates the memory space
needed for a parameter vector and returns a pointer to it.
\end{function}
\begin{function}
\GLBNS{AllocParams}
{\tt void glbFreeParams(glb\_params stale)} destroys the memory space
needed for a parameter vector {\tt stale} and sets the pointer to {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{DefineParams}
{\tt glb\_params glbDefineParams(glb\_params in, double theta12, double theta13,double theta23, double delta, double dms, double dma)} assigns
the complete set of oscillation parameters to the vector {\tt in}, which has
to be allocated before. The return value is the pointer to {\tt in} if 
the assignment was successful, and {\tt NULL} otherwise.
\end{function}
\begin{function}
 \GLBNS{CopyParams}
 {\tt glb\_params glbCopyParams(const glb\_params source, glb\_params dest)}
 copies the vector {\tt source} to the vector {\tt destination}. The return
 value is {\tt NULL} if the assignment was not successful.
\end{function}
\begin{function}
\GLBNS{PrintParams}
{\tt void glbPrintParams(FILE *stream, const glb\_params in)} prints the
parameters in {\tt in} to the file {\tt stream}. The oscillation
parameters, all density values, and the number of iterations are
printed as pretty output. Use {\tt stdout} for {\tt stream} if you want
to print to the screen.
\end{function}
In addition to these basic functions, there are functions to access
the individual parameters within the parameter vectors:
\begin{function}
\GLBNS{SetOscParams}
{\tt glb\_params glbSetOscParams(glb\_params in, double osc, int which)}
sets the oscillation parameter {\tt which} in the structure {\tt in}
to the value {\tt osc}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{GetOscParams}
{\tt double glbGetOscParams(glb\_params in, int which)}
returns the value of the oscillation parameter {\tt which} in the 
structure {\tt in}.
\end{function}
In both of these functions, the parameter {\tt which} runs from
$0$ to $5$, where the parameters in \GLOBES\ always have the order $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$.
Alternatively to the number, the constants \GLBC{GLB\_THETA\_12}, \GLBC{GLB\_THETA\_13}, \GLBC{GLB\_THETA\_23}, \GLBC{GLB\_DELTA\_CP}, 
\GLBC{GLB\_DM\_SOL}, or \GLBC{GLB\_DM\_ATM} can be used.

Similarly, the density parameters or number of iterations 
(returned by the minimizers) can be accessed:
\begin{function}
\GLBNS{SetDensityParams}
{\tt glb\_params glbSetDensityParams(glb\_params in, double dens, int which)} sets the density parameter {\tt which} in the structure {\tt in}
to the value {\tt dens}. If the assignment was unsuccessful, the 
function returns {\tt NULL}. If {\tt GLB\_ALL} is used for {\tt which},
the density parameters of all experiments will be set accordingly.
\end{function}
\begin{function}
\GLBNS{GetDensityParams}
{\tt double glbGetDensityParams(glb\_params in, int which)}
returns the value of the density parameter {\tt which} in the 
structure {\tt in}.
\end{function} 
\begin{function}
\GLBNS{SetIteration} \index{norm}{Iterations}
{\tt glb\_params glbSetIteration(glb\_params in, int iter)} sets the number of iterations in the structure {\tt in}
to the value {\tt iter}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{GetIteration}
{\tt int glbGetIteration(glb\_params in)}
returns the value of the number of iterations in the 
structure {\tt in}.
\end{function}
In total, the parameter vector handling in a program normally has the
following order:\index{norm}{Parameter vector handling}
\begin{quote}
{\tt
glbInitExperiment(...); \\
/* ... more initializations ... */ \\
\\
glb\_params vector1 = glbAllocParams(); \\
/* ... more vectors allocated ... */ \\
\\
/* Program code: assign and use vectors */ \\
\\
glbFreeParams(vector1); \\
/* ... more vectors destroyed ... */ \\
\\
/* ... end of program or glbClearExperimentList ... */ 
}
\end{quote}

@@

Compared to existing experiments, which use real data, future experiments
uses simulated data. Thus, the {\em true parameter values} and their results in form of the reference event rate vectors are simulated. After setting the true parameter values, the {\em fit parameter values} can be varied in order to obtain information on the measurement performance for the given set of true parameter values. Therefore, it is often useful to show the results of a future measurement as function of the true parameter values for which the reference rate vectors are computed -- at least within the currently allowed ranges. The true parameter values for the vacuum neutrino oscillation parameters have to be set by the functions \GLB{SetOscillationParameters} and \GLB{SetRates} {\em before} any evaluation function is used and {\em after} the experiments have been initialized and the experiment parameters have been adjusted which could change the rates (such as baseline or target mass). This means that
after any change of an experiment parameter, {\tt glbSetRates} has to be called. Matter effects are automatically included as specified in the experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{SetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{GetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{SetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

The functions in \GLOBES\ for version control are:
\begin{function}
\GLBNS{TestReleaseVersion}
{\tt int glbTestReleaseVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer. 
\end{function}
\begin{function}
\GLBNS{TestLibraryVersion}
{\tt int glbTestLibraryVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer.
Note that the library and \GLOBES\ versions are not the same. 
\end{function}
\begin{function}
\GLBNS{VersionOfExperiment} 
{\tt const char* glbVersionOfExperiment(int experiment)} returns
the version string of the experiment number {\tt experiment}.
The version string is allocated within the experiment structure, which means
that it cannot be altered and must not be destroyed by the
user.
\end{function}

\chapter[Calculating $\chi^2$ with systematics only]{Calculating $\boldsymbol{\chi^2}$ with systematics only}
\index{norm}{Systematics $\chi^2$}

Calculating a $\chi^2$-value with or without systematics, but no correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. In general, \GLOBES\ uses the six independent oscillation parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$, as well as the matter density 
scaling factor $\hat{\rho}$ of each experiment.\index{norm}{Matter density
scaling factor} Thus, there are six plus the number of experiments parameters determining the rate vectors. Using the matter density scaling factors  in addition to the oscillation parameters will allow the simulation of the correlations with matter density uncertainties: In this
approach, the matter density profile normalization $\hat{\rho}$ can 
be treated
as parameter to be measured by the experiment, where an external precision
given by observations is imposed (typically up to $5\%$). 
\index{norm}{Matter density uncertainty}
 For this section, it is important to keep in mind
that there are more parameters than just the oscillation parameters
determining the simple $\chi^2$. However, as we have described in the
last section, the mechanism for the matter density scaling factors
is hidden in the definition of \GLB{\_params}: Each of the scaling
factors is initially set to $1.0$. Therefore, for the calculation of
$\chi^2$ with systematics only, we do not have to care about the
matter density scaling factors.

@@

\chapter[Calculating $\chi^2$-projections: how one can include correlations]{Calculating $\boldsymbol{\chi^2}$-projections: how one can include correlations}
\label{chapt:correlations}
\index{norm}{Correlations $\chi^2$}

\index{norm}{Multi-parameter correlation}
This chapter deals with the rather complicated issue of $n$-parameter correlations. It is one of the greatest strengths of this software 
to include the full $n$-parameter correlation in the high-dimensional parameter space with reasonable effort. Of course, calculating $\chi^2$-projections is somewhat more complicated than using systematics only. Therefore, we use a simple step by step introduction to the problem.

@@

\section[Projection onto any hyperplane]{Projection onto any hyperplane}
\index{norm}{Projection onto hyperplane}

>>>>

In this first chapter, we show a \GLOBES\ tour illustrating the
main features of \GLOBES . The complete example  
can be found as {\tt example-tour.c} in the \verb+example+ subdirectory 
of your \GLOBES\ distribution.
The output is written to {\tt stream}, which can be either {\tt stdout},
or a file. Details about how to use \GLOBES\ with C can found in \Chapt~\ref{chapt:gettingstarted} and the following chapters.
You can also find a summary of the most important \GLOBES\ $\chi^2$-functions in \tabl{stdfunctions}. Note that this chapter
can be skipped without loss of relevant information.

@@

 for(i=1;i<4;i++) fprintf(stream,"1->\%i: \%g ",i,\\ \hspace*{4cm} glbProfileProbability(0,1,i,+1,50)); \\

@@

we have to free them first:

@@

Finally, we have to free the parameter vectors again:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\GLOBES\ basics}
\label{chapt:gettingstarted}

\index{norm}{Installation}
In this first chapter of the user's manual, we assume that the \GLOBES\ 
software is readily installed on your computer system. For the installation,
see \App~\ref{app:installation} and the {\tt INSTALL} file in the
software package. We demonstrate how to load pre-defined experiments 
and introduce the basic concepts of \GLOBES . We do not go
into details of the programming language, which means that standard parts
of the program code common to all of the examples in the following chapters 
are, in general, omitted.
An example of a minimal \GLOBES\ program in C can be found on 
page~\pageref{ex:c}. Furthermore, the files of the examples in this part 
can be found in the \verb+example+ subdirectory of your \GLOBES\ distribution.
 \index{norm}{Examples} 
After the installation of \GLOBES, they can be compiled using 
the \verb+Makefile+ in the \verb+examples+
 directory. The  \verb+Makefile+ has been correctly setup by the 
\verb+configure+
script to take into account details of the installation on your system. Thus
you've just to type \verb+make+ and you're done.
\footnote{The  data files (\AEDL\ and supporting files) needed by the
examples are already in place.} This \verb+Makefile+ very well serves
as a template for your own applications.
\index{norm}{Compilation!of application programs}
\index{norm}{Build process| \see{Compilation}}

@@

\hspace*{0.5cm}  /* Free parameter vector(s) */ \\
\hspace*{0.5cm}  glbFreeParams(true\_values); \\
\hspace*{0.5cm} /* ... */ \\

@@

\index{norm}{Experiment!number of}
In principle, the \GLOBES\ user interface can currently handle up to 32 of different long-baseline experiments simultaneously, where the number
of existing experiment definition files can, of course, be unlimited. This means that their $\Delta \chi^2$-values are added {\em after} the minimization over the systematics parameters, and {\em before} any minimization over the oscillation parameters. Note that each experiment
assumes a specific matter density profile, which means
that it makes sense to simulate different operation modes within one
experiment definition, and physically different baselines in different
definitions. For details of the rate computation and
simulation techniques, we refer at this place to \Part~\ref{part:2}. Though
 the simplest case of simulating one experiment may be most often used, 
 using more than one experiments are useful in many cases. For example, combinations of experiments can be tested for
complementarity and competitiveness by equal means within one program.
In general, many \GLOBES\ functions take the experiment number as
a parameter, which runs from $0$ to \GLB{\_num\_of\_exps}-1 in the order of their initialization in the program.\footnote{Note that
the global variable {\tt glb\_num\_of\_exps} must not be modified by the
user.} In addition, using the parameter value \GLBC{GLB\_ALL} as
experiment number initiates a combined analysis of all loaded experiments.
\index{norm}{GLB ALL@{\tt GLB\_ALL}}
\index{constants}{{\tt GLB\_ALL}}

In general \GLBC{GLB\_ALL} can be used in many cases where there is an argument
selecting `$i$ out of $N$', \eg\ the 1st experiment out of 5, or the 5th
rule of 20. In those cases  using \GLBC{GLB\_ALL} is equivalent to
calling the corresponding function for \emph{all} $i$ in $N$  and `add' 
the effect of each
invocation, like in
\begin{quote}
\verb&for (i=0;i<N;i++) result += some_function(i);&\\
is the same as\\
\verb&result = some_function(GLB_ALL);&
\end{quote}
Here the meaning of `add' is, that whatever the desired result of calling
\verb+some_function+ is, this result is obtained for each $i$ in $N$, \eg\ 
setting the baseline in all experiments to a certain value or
or compute the $\chi^2$ for each experiment and return the total result.
There are however some functions where the action performed or the result
is so complex that is not possible or sensible to perform this for all $i$
in $N$. Calling these functions with  \GLBC{GLB\_ALL} as argument will 
in any case result in an exit status indicating
failure and the function will produce an 
error message\footnote{if the verbosity
level is set accordingly}.

\index{norm}{Experiment!list}
\index{norm}{Initialization!experiments}
For storing the experiments, \GLOBES\ uses the initially empty list of experiments \GLB{\_experiment\_list}. To add a pre-defined experiment to this list, one can use the function {\tt glbInitExperiment}:
\begin{function}
\index{norm}{Experiment initialization}
\GLBNS{InitExperiment}
{\tt int glbInitExperiment(char *inf, glb\_exp *in, int *counter)}
 adds a single experiment with the filename {\tt inf} to the list of currently loaded experiments. The {\tt counter} is a pointer to the 
 variable containing the number of experiments, and the experiment {\tt in}
 points to the beginning of the experiment list. The function returns
 zero if it was successful. 
\end{function}
Normally, a typical  call of {\tt glbInitExperiment} is 
\begin{quote}
{\tt glbInitExperiment("NuFact.glb",\&glb\_experiment\_list[0],\\  \hspace*{8cm} \&glb\_num\_of\_exps); }
\end{quote}
In this case, the experiment in the file {\tt NuFact.glb} is added to the internal global list of experiments, and the experiment counter is increased. The
experiment then has the number {\tt glb\_num\_of\_exps}-1. The elements
of the experiment list have the type \GLB{\_exp}, which the
user will not need to access directly in most cases. The experiment definition files, which usually end with {\tt .glb}, and any
supporting files, are first of all searched in the current directory, and then in the path given in the environment variable \verb+GLB_PATH+.
%
\index{norm}{GLB PATH@{\tt GLB\_PATH}}
\index{norm}{Environment variables!GLB PATH@{\tt GLB\_PATH}}
%

\index{norm}{Path resolution}
A list of pre-defined experiment prototypes, their filenames, their short descriptions, and the references of their definitions can be found in \tabl{experiments}. If the program cannot find these files, or some of them are syntactically not correct, it will break at this place.

@@

Before we can set the simulated event rates or access any oscillation
parameters, we need to become familiar with the concept \GLOBES\ uses
for oscillation parameters. In order to transfer sets of oscillation
parameter vectors $(\theta_{12}, \theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm)$, the parameter type \GLB{\_params} is used. In general,
 this type is often
transferred to and from \GLOBES\ functions. Therefore, the memory 
for these vectors has to be reserved (allocated) before they can be used, 
and it has
to be returned (freed) afterwards. \GLOBES\ functions usually use the pointers
of the type {\tt glb\_params} for the input or output to the functions. As
an input parameter, the pointer has to be created and point towards a valid
parameter structure, where the oscillation parameters are read from. As
an output parameter, the pointer has to be created, too, and point towards
a structure which will contain the return values will be written to.
This parameter transfer concept seems to be very sophisticated, but, as we
will see in the next chapters, it hides a lot of complicated parameter 
mappings which otherwise need to be done by the user. For example, not
only the oscillation parameters are stored in the pointer structure, but also information on the matter
densities of all of the initialized experiments. Since \GLOBES\ treats
the matter density as a free parameter known with some external precision
to include matter density uncertainties,
the minimizers also use fit values and external errors for the matter
densities of all loaded experiments. More precisely, the matter density
profile of each experiment $i$ is multiplied by a scaling factor
$\hat{\rho_i}$, \index{norm}{Matter density!scaling factor}
\index{norm}{Matter density!profile} 
which is stored in the density information of {\tt glb\_params}. 
Each of
these scaling factors has $1.0$ as pre-defined value. Since it
is in most cases not necessary to change this value, the user does
not need to take care of it. For a constant matter density, it is simply the ratio of the matter density and the average matter density specified in the experiment definition, \ie , $\hat{\rho_i} \equiv \rho_i/\bar{\rho}_i$. For a matter density profile, it acts as an overall normalization factor: The matter density in each layer is multiplied by this factor. In most cases one wants to take a scaling factor of $1.0$ here, which simply means taking the matter density profile as it is given in the experiment definition. For the treatment of correlations, however,
an external precision of the scaling factor might be used to include the
correlations with the matter density uncertainty.
Note that the {\tt glb\_params}
structures must not be initialized before all experiments are loaded,
since the number of matter densities can only be determined after the
experiments are initialized. 
Similarly, any change in the number of experiments requires that the
parameter structures be re-initialized, \ie, freed and allocated again.

Another piece of information will be returned from the minimizers 
(\cf, \Chapt~\ref{chapt:correlations}) and transferred
into the {\tt glb\_params} structure is the number
of iterations used for the minimization, which is proportional to the
running time of the minimizer. In general, the user does not need to
access the elements in {\tt glb\_params} directly. A number of functions
is provided to handle these parameter structures:
\begin{function}
\GLBNS{AllocParams}
{\tt glb\_params glbAllocParams()} allocates the memory space
needed for a parameter vector and returns a pointer to it.
\end{function}
\begin{function}
\GLBNS{AllocParams}
{\tt void glbFreeParams(glb\_params stale)} frees the memory
needed for a parameter vector {\tt stale} and sets the pointer to {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{DefineParams}
{\tt glb\_params glbDefineParams(glb\_params in, double theta12, double theta13,double theta23, double delta, double dms, double dma)} assigns
the complete set of oscillation parameters to the vector {\tt in}, which has
to be allocated before. The return value is the pointer to {\tt in} if 
the assignment was successful, and {\tt NULL} otherwise.
\end{function}
\begin{function}
 \GLBNS{CopyParams}
 {\tt glb\_params glbCopyParams(const glb\_params source, glb\_params dest)}
 copies the vector {\tt source} to the vector {\tt destination}. The return
 value is {\tt NULL} if the assignment was not successful.
\end{function}
\begin{function}
\GLBNS{PrintParams}
{\tt void glbPrintParams(FILE *stream, const glb\_params in)} prints the
parameters in {\tt in} to the file {\tt stream}. The oscillation
parameters, all density values, and the number of iterations are
printed as pretty output. Use {\tt stdout} for {\tt stream} if you want
to print to the screen.
\end{function}
In addition to these basic functions, there are functions to access
the individual parameters within the parameter vectors:
\begin{function}
\GLBNS{SetOscParams}
{\tt glb\_params glbSetOscParams(glb\_params in, double osc, int which)}
sets the oscillation parameter {\tt which} in the structure {\tt in}
to the value {\tt osc}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{GetOscParams}
{\tt double glbGetOscParams(glb\_params in, int which)}
returns the value of the oscillation parameter {\tt which} in the 
structure {\tt in}.
\end{function}
In both of these functions, the parameter {\tt which} runs from
$0$ to $5$, where the parameters in \GLOBES\ always have the order $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$.
Alternatively to the number, the constants \GLBC{GLB\_THETA\_12}, \GLBC{GLB\_THETA\_13}, \GLBC{GLB\_THETA\_23}, \GLBC{GLB\_DELTA\_CP}, 
\GLBC{GLB\_DM\_SOL}, or \GLBC{GLB\_DM\_ATM} can be used.

Similarly, the density parameters or number of iterations 
(returned by the minimizers) can be accessed:
\begin{function}
\GLBNS{SetDensityParams}
{\tt glb\_params glbSetDensityParams(glb\_params in, double dens, int which)} sets the density parameter {\tt which} in the structure {\tt in}
to the value {\tt dens}. If the assignment was unsuccessful, the 
function returns {\tt NULL}. If {\tt GLB\_ALL} is used for {\tt which},
the density parameters of all experiments will be set accordingly.
\end{function}
\begin{function}
\GLBNS{GetDensityParams}
{\tt double glbGetDensityParams(glb\_params in, int which)}
returns the value of the density parameter {\tt which} in the 
structure {\tt in}.
\end{function} 
\begin{function}
\GLBNS{SetIteration} \index{norm}{Iterations}
{\tt glb\_params glbSetIteration(glb\_params in, int iter)} sets the number of iterations in the structure {\tt in}
to the value {\tt iter}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{GetIteration}
{\tt int glbGetIteration(glb\_params in)}
returns the value of the number of iterations in the 
structure {\tt in}.
\end{function}
In total, the parameter vector handling in a program normally has the
following order:\index{norm}{Parameter vector handling}
\begin{quote}
{\tt
glbInitExperiment(...); \\
/* ... more initializations ... */ \\
\\
glb\_params vector1 = glbAllocParams(); \\
/* ... more vectors allocated ... */ \\
\\
/* Program code: assign and use vectors */ \\
\\
glbFreeParams(vector1); \\
/* ... more vectors freed ... */ \\
\\
/* ... end of program or glbClearExperimentList ... */ 
}
\end{quote}

@@

Compared to existing experiments, which use real data, future experiments
use simulated data. Thus, the {\em true parameter values} and their results in 
form of the reference event rate vectors are simulated. After setting the 
true parameter values, the {\em fit parameter values} can be varied in order to 
obtain information on the measurement performance for the given set of true 
parameter values. Therefore, it is often useful to show the results of a 
future measurement as function of the true parameter values for which the 
reference rate vectors are computed -- at least within the currently allowed 
ranges. The true parameter values for the vacuum neutrino oscillation 
parameters have to be set by the function \GLB{SetOscillationParameters} 
and the reference rate vector, \ie\ the data, has to be computed by a call 
to \GLB{SetRates}. This has to be done {\em before} any evaluation 
function is used and  {\em after} the experiments have been initialized and 
also the experiment 
parameters have been adjusted which could change the rates (such as baseline
 or target mass). This means that
after any change of an experiment parameter, \GLB{SetRates} has to be 
called. Matter effects are automatically included as specified in the 
experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{SetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{GetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{SetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

The functions in \GLOBES\ for version control are:
\begin{function}
\GLBNS{TestReleaseVersion}
{\tt int glbTestReleaseVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer. 
\end{function}
\begin{function}
\GLBNS{TestLibraryVersion}
{\tt int glbTestLibraryVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer.
Note that the library and \GLOBES\ versions are not the same. 
\end{function}
\begin{function}
\GLBNS{VersionOfExperiment} 
{\tt const char* glbVersionOfExperiment(int experiment)} returns
the version string of the experiment number {\tt experiment}.
The version string is allocated within the experiment structure, which means
that it cannot be altered and must not be freed by the
user.
\end{function}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter[Calculating $\chi^2$ with systematics only]{Calculating $\boldsymbol{\chi^2}$ with systematics only}
\index{norm}{Systematics $\chi^2$}

Calculating a $\chi^2$-value with or without systematics, but no correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. In general, \GLOBES\ uses the six independent oscillation parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$, as well as the matter density 
scaling factor $\hat{\rho}$ of each experiment.\index{norm}{Matter density!
scaling factor} Thus, there are six plus the number of experiments parameters determining the rate vectors. Using the matter density scaling factors  in addition to the oscillation parameters will allow the simulation of the correlations with matter density uncertainties: In this
approach, the matter density profile normalization $\hat{\rho}$ can 
be treated
as parameter to be measured by the experiment, where an external precision
given by observations is imposed (typically up to $5\%$). 
\index{norm}{Matter density!uncertainty}
 For this section, it is important to keep in mind
that there are more parameters than just the oscillation parameters
determining the simple $\chi^2$. However, as we have described in the
last section, the mechanism for the matter density scaling factors
is hidden in the definition of \GLB{\_params}: Each of the scaling
factors is initially set to $1.0$. Therefore, for the calculation of
$\chi^2$ with systematics only, we do not have to care about the
matter density scaling factors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter[Calculating $\chi^2$-projections: how one can include correlations]{Calculating $\boldsymbol{\chi^2}$-projections: how one can include correlations}
\label{chapt:correlations}
\index{norm}{Correlations $\chi^2$}

\index{norm}{Correlations!multi-parameter}
This chapter deals with the rather complicated issue of $n$-parameter correlations. It is one of the greatest strengths of this software 
to include the full $n$-parameter correlation in the high-dimensional parameter space with reasonable effort. Of course, calculating $\chi^2$-projections is somewhat more complicated than using systematics only. Therefore, we use a simple step by step introduction to the problem.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Projection onto any hyperplane]{Projection onto any hyperplane}
\index{norm}{Projection!hyperplane}
